Initial environmental impacts of Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-08-11 manu-draft.Rmd
intro draft

From March 30th to April 9th, 2021, 215 million gallons of legacy phosphate mining wastewater from the Piney Point facility were released into Tampa Bay (Florida, USA).  An estimated ~205 tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual nitrogen load estimates in a matter of days.  An initial phytoplankton response was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L.  Macroalgae blooms of cyanobacteria (*Lyngbya* spp.) were observed in May, with biomass estimated at 0.5 kg/m2 at some locations.  Blooms of *Karenia brevis* were observed beginning in May and continuing through July.  Reported fish kills tracked bloom concentrations, prompting local cleanup efforts to remove over 1700 tons of dead fish.  Combined, these observations indicate abnormal conditions in Tampa Bay following release of wastewater from Piney Point, which is supported by comparison to the decades of baseline environmental monitoring data for the region.

@@

* Nutrient management paradigm for southwest Florida estuaries, historical progress
* Threats/challenges to protecting water quality: nutrient inputs, climate change
* Wastewater byproducts from mining are a global threat to the quality of surface and groundwater sources worldwide [@Hudson11]. 
     * [@Tayibi09] reviews environmental impact of phosphogypsum production, including contaminantion of groundwater/leaching to coastal waters.  
     * [@Burnett01] evaluate radiochemistry of phosphogypsum at Piney Point to evaluate potential for contamination of groundwater sources [pdf](https://www.osti.gov/etdeweb/servlets/purl/20243518)
    * Citations from [@Perez16] to track down: Santos-Cubatao estuarine system (Brazil; Sanders et al., 2013), in the Lebanese coast (El Samad et al., 2014) and in the Gulf of Gabes (Tunisia; El Zrelli et al., 2016)
    * Dillon report about Grand Bay [link](https://www.wrri.msstate.edu/pdf/2016dillon_finalreport.pdf)
* Phosphate mining industry in Florida, then/now
* Piney Point
* Goals/objectives

This paper provides an initial assessment of environmental conditions in Tampa Bay following the release of 215 million gallons of legacy phosphate mining wastewater over a ten-day period.  The goal is to describe the results of monitoring data collected in response to the discharge event to assess deviation of current conditions relative to long-term, seasonal records of water quality, phytoplankton, seagrass, benthic, and nekton datasets available for the region.  We provide a brief overview of the history of the Piney Point facility, including past wastewater releases and impacts observed in Tampa Bay.  A timeline of events in 2021 is also provided, which is supported by the results from response-based monitoring of conditions in and around Piney Point.  The results of this study provide an initial documentation of impacts to the natural resources of Tampa Bay that can be used to inform long-term assessments of acute wastewater discharge events on the environmental quality of the region.

@@

* Comparison to other locations/past events - Grand Bay, Bishop Harbor, Huelva estuary [@Perez10;@Perez16]
* Analysis limitations: no smoking gun but 2021 is an anomaly, additional info (benthic diversity TBD, nekton diversity TBD, large mammals, etc.)
* Potential long-term impacts TBD
* Current challenges in TB/southwest FL - OTB, seagrass loss, red tide, climate change
* Risk of decline (IRL ex.), regression of past progress

>>>>

From March 30th to April 9th, 2021, 215 million gallons of legacy phosphate mining wastewater from the Piney Point facility were released into Tampa Bay (Florida, USA).  An estimated 205 tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual nitrogen load estimates in a matter of days.  An initial phytoplankton response was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L. Macroalgae blooms of cyanobacteria (*Lyngbya* spp.) were observed in May, with biomass estimated at 0.5 kg/m2 at some locations.  Blooms of *Karenia brevis* were observed beginning in May and continuing through July.  Reported fish kills tracked bloom concentrations, prompting local cleanup efforts to remove over 1700 tons of dead fish.  Combined, these observations indicate abnormal conditions in Tampa Bay following release of wastewater from Piney Point, which is supported by comparison to the decades of baseline environmental monitoring data for the region.

@@

Ecosystem management paradigms for estuaries of the Gulf Coast of Florida, USA are based primarily on the control of nutrient pollutants from stormwater and wastewater sources.  The effects of nitrogen from source inputs are well understood as a limiting nutrient for the growth of algal blooms that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14]. Seagrasses in particular are a primary endpoint for assessing the impacts of nutrient pollution on water quality based on established relationships between nitrogen, phytoplankton growth, water clarity, and light requirements for seagrass species observed in nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18g].  Tampa Bay is the largest estuary in Florida located in a heavily urbanized watershed of nearly 3 million individuals.  Historical gains in seagrass coverage in Tampa Bay have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads into major bay segments [@Janicki96;@Greening16].  Together, these efforts have resulted in the long-term recovery of Tampa Bay through a reduction in nitrogen loads, improvements in water clarity, and baywide expansion of seagrass coverage to benchmark targets established for the region [@Greening14;@Sherwood17].

Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite historical gains in environmental recovery.  Although point-source inputs of nutrient loads from wastewater treatment plants into Tampa Bay have been reduced, non-point sources from wastewater and stormwater runoff contribute to nutrient inputs, particularly during the rainy season from June to September [@Janicki08;@Janicki17].  Atmospheric deposition of nutrients from coal-based power production and automobile traffic further contribute about one-quarter of the total nitrogen inputs to the bay [@Poor13].  Climate change stressors, such as sea level rise, changing rainfall patterns, and temperature alterations, may further perturb ecosystem dynamics and assimilative capacity by reducing system resilience to nutrient inputs [@Sherwood14;@Burke17].  Many of these challenges are addressed by ongoing efforts of the US EPA National Estuary Program to implement a science-based resource management plan for the Bay[@Ohara17].  The Tampa Bay Estuary Program has been instrumental in coordinating efforts among local and regional partners to address legacy pollutants and current threats to the long-term protection of bay resources [@Greening14;@Greening16].

Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources worldwide [@Tayibi09;@Hudson11]. Fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].  The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of waste precipitate, or phosphogypsum (CaSO$_4\cdot$ H$_2$O).  Impurities, contaminants, and radionuclides exist in phosphogypsum, making it commercially invaluable and the resulting waste is typically stored on-site in large earthen stacks (gypstacks) or holding ponds [@Burnett01].  The stacks are usually near distribution centers where fertilizer is shipped elsewhere, such as port facilities close to coastal resources or population centers [@Beck18]. There are obvious environmental and human health risks associated with these stacks, primarily through controlled or uncontrolled discharge to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks. Examples exist worldwide demonstrating the potential harm of these facilities on the environment [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18].

The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer used in food production [@Henderson04].  By 2001, an estimated 40 million tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].  Currently, fourteen phosphogypsum stacks exist in the Tampa Bay watershed with no long-term plan for closure or disposal of waste to preven impacts to the environment.  The Piney Point facility located in Palmetto, Florida is a large phosphogypsum stack established in 2001 located less than two miles from the shore of Tampa Bay and near two aquatic preserves [@Henderson04].  Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, with oversight by the Florida Departmental of Environmental Protection (FLDEP).  Decreasing holding capacity of the ponds with seasonal rain events and storage of dredging material from nearby Port Manatee have contributed to degradation of the facility.  Discharges of wastewater from the stacks occurred in 2010 to nearby Bishops Harbor connected to Tampa Bay [@Garrett11].  Recently, FLDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 29th, 2021 to release wastewater from the stacks directly into lower Tampa Bay to prevent catastrophic failure of the berms supporting the holding ponds.

This paper provides an initial assessment of environmental conditions in Tampa Bay over five months following the recent release of 215 million gallons of legacy phosphate mining wastewater in April, 2021.  The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess deviation of current conditions relative to long-term, seasonal records of water quality, phytoplankton, seagrass, benthic, and nekton datasets available for the region.  We provide a brief overview of the history of the Piney Point facility, including past wastewater releases and impacts observed in Tampa Bay.  A timeline of events in 2021 is also provided, which is supported by the results from 2021 response-based monitoring of conditions in and around Piney Point.  The results of this study provide an initial documentation of impacts to the natural resources of Tampa Bay that can be used to inform long-term assessments of acute wastewater discharge events on the environmental quality of the region.  We focus primarily on the perspective of the Tampa Bay Estuary Program in its role in coordinating monitoring and evaluating short-term impacts, particularly in the context of long-term management goals that leverage existing partnerships among local resource management institutions.

@@

* Comparison to other locations/past events - Grand Bay, Bishop Harbor, Huelva estuary [@Perez10;@Perez16], Dillon report about Grand Bay [link](https://www.wrri.msstate.edu/pdf/2016dillon_finalreport.pdf)
* Analysis limitations: no smoking gun but 2021 is an anomaly, additional info (benthic diversity TBD, nekton diversity TBD, large mammals, etc.), response-based monitoring may be biased
* Potential long-term impacts TBD
* Current challenges in TB/southwest FL - OTB, seagrass loss, red tide, climate change
* Risk of decline (IRL ex.), regression of past progress
